---
title: "Reproducible Research Week 2 Course Project 1"


output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(ggplot2)
library(dplyr)

```

## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date)
```
## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day.
```{r}
stepsPerDay <- activity %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
#Display first 10 rows of data
head(stepsPerDay,10)
```
2. Make a histogram of the total number of steps taken each day.
```{r}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="yellow", xlab="Steps", ylim = c(0,30))
```
3. Calculate and report the mean and median of the total number of steps take per day.
```{r}
meanPreNA <- round(mean(stepsPerDay$sumsteps),digits = 2)
medianPreNA <- round(median(stepsPerDay$sumsteps),digits = 2)

print(paste("The mean is: ", meanPreNA))

```

```{r}
print(paste("The median is: ", medianPreNA))
```
## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = “l”|) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r}
stepsPerInterval <- activity %>%
        group_by(interval) %>%
        summarize(meansteps = mean(steps, na.rm = TRUE)) 
#Display first 10 rows of data
head(stepsPerInterval,10)
```

```{r}
plot(stepsPerInterval$meansteps ~ stepsPerInterval$interval,
     col="green", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
print(paste("Interval containing the most steps on average: ",stepsPerInterval$interval[which.max(stepsPerInterval$meansteps)]))
```

```{r}
print(paste("Average steps for that interval: ",round(max(stepsPerInterval$meansteps),digits=2)))
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps))))
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
#Before 
#Display first 10 rows of data
head(activity,10)
```

```{r}
activityNoNA <- activity  
for (i in 1:nrow(activity)){
        if(is.na(activity$steps[i])){
                activityNoNA$steps[i]<- stepsPerInterval$meansteps[activityNoNA$interval[i] == stepsPerInterval$interval]
        }
}

#After
#Display first 10 rows of data
head(activityNoNA,10)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
stepsPerDay <- activityNoNA %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsPerDay,10)
```

```{r}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="orange", xlab="Steps")
```

```{r}
meanPostNA <- round(mean(stepsPerDay$sumsteps), digits = 2)
medianPostNA <- round(median(stepsPerDay$sumsteps), digits = 2)

print(paste("The mean is: ", mean(meanPostNA)))
```

```{r}
print(paste("The median is: ", median(medianPostNA)))
```

```{r}
NACompare <- data.frame(mean = c(meanPreNA,meanPostNA),median = c(medianPreNA,medianPostNA))
rownames(NACompare) <- c("Pre NA Transformation", "Post NA Transformation")
print(NACompare)
```


##Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
activityDoW <- activityNoNA
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)

```

2. Make a panel plot containing a time series plot (i.e. type=“l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

activityWeekday <- activityWeekday %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekday$day <- "weekday"

activityWeekend <- activityWeekend %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekend$day <- "weekend"

wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)


g <- ggplot (wkdayWkend, aes (interval, steps))
g + geom_line() + facet_grid (day~.) + 
        theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
        labs(y = "Number of Steps") + labs(x = "Interval") + 
        ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
        theme(plot.title = element_text(hjust = 0.5))
```

